Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-04-01 paper.tex
revising, added soft intro

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disks such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. % with, for example, radial variations in temperature. 
  However, the VSI requires a short thermal or cooling 
  timescale in sub-adabatically (stably) stratified disks.  
  A quantitative examination of this thermodynamic 
  requirement is needed to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of stably stratified accretion disks in which the disk  
  temperature $T$ is relaxed to its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response of the disk by
  varying $\beta$. We find that 
  increasing $\beta$ even slightly away from    
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  is the radial temperature gradient, and 
  $\gamma$ is the adiabatic index. 
  We apply our results  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. In the inner parts of the MMSN, e.g. at $\lesssim 5$AU, 
  we find the VSI would only operate on extremely short radial lengthscales, $\ll 
  O(10^{-2}H)$, where $H$ is the disk scale height. On longer radial 
  lengthscales of $O(10^{-2}H)$ to $O(10^{-1}H)$, we find the MMSN is
  unstable to the VSI from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.  
  % ,and the dominant VSI mode in the MMSN is the `fundamental
  % corrugation mode' which causes the disk column to oscillate
  % vertically.
  We thus suggest the VSI is at least dynamically important in the 
  outer parts of protoplanetary disks at tens of AU. 
\end{abstract}

\section{Introduction}
Understanding the mechanisms of angular momentum transport is the
central problem in the theory of accretion disks 
\citep{lyndenbell74}. The   
magneto-rotational instability in the context of differentially
rotating disks \citep[MRI,][]{balbus91} is now accepted as a robust
pathway to magneto-hydrodynamic (MHD) turbulence, which transports
angular momentum outwards and thus enables mass accretion.

@@

The VSI can be regarded as the disk analog of the
Goldreich-Schubert-Fricke instability 
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
A height-dependence in $\Omega$, or vertical shear, can render
inertial waves unstable. \cite{barker15} gives a simple 
interpretation of the VSI: in an axisymmetric inviscid disk, the 
exchange of fluid parcels vertically, while conserving their specific angular
momentum, can result in a new configuration of lower energy if the
distribution of specific angular momentum varies with $z$. The VSI is
characterized by perturbations with large vertical length-scales 
relative to radial length-scales.

Early analysis of the VSI in radially and vertically local 
models of accretion disks \citep{urpin98,urpin03} showed that the VSI
requires a short thermal timescale so that perturbations evolve 
isothermally, unless the disk is neutrally stratified in the vertical
direction. This is because
vertical motions associated with the VSI 
are stabilized by buoyancy forces if the disk is sub-adiabatically
stratified (convectively stable). A sufficiently short thermal
timescale renders buoyancy ineffective.    
%efficient exchang of heat with surroundings

@@

These studies show that the VSI is generally sensitive to disk
thermodynamics. It is thus important to quantify this dependence, for
example on disk parameters, in order to evaluate the relevance of the
VSI to realistic disks, which one may expect to be stably
stratified and have non-zero thermal timescales. For example, we show in  
Fig. \ref{bcrit_mmsn2d} an estimate of the thermal timescales in a
typical protoplanetary disk model (see below). In order to determine
where in the disk the VSI may operate, and hence lead to hydrodynamic
turbulence, understanding the thermodynamic dependence of the linear VSI is essential. 
% since the transition
% Should a disk
% become turbulent through the VSI, it must first pass through the linear
% instability.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/bcrit_mmsn2d} 
  \caption{Estimate of the thermal timescale  in a fiducial 
    protoplanetary disk model, the Minimum Mass Solar Nebula, for
    perturbations with radial lengthscales $\sim 10^{-2}H$, where $H$
    is the disk scale-height. The red line corresponds to a 
    characteristic thermal timescale, derived in this work, 
    below which the VSI is expected to operate effectively (the region
    outside the line).   
    \label{bcrit_mmsn2d} 
  }
\end{figure}

@@

This paper is organized as follows. In \S\ref{setup} we list the
governing equations and global equilibrium disk models under
consideration. We formulate the linear stability problem in
\S\ref{linear} and simplify it by making the radially local
approximation. We study the linear problem using two approaches. In
\S\ref{analytical} we present an analytical discussion of vertically
isothermal disks. Our treatment is grossly simplified owing to a
number of approximations made, but closely parallels previous work and 
allows us to highlight the essential effects of vertical shear. In
particular, we quantify a characteristic upper limit to the thermal
timescale for the VSI. In \S\ref{numerical} we numerically
solve the linear problem in the radially local approximation, but
relax all other approximations made in our analytical approach.    
We find basic agreement with our analysis, including
the aforementioned characteristic thermal timescale. In
\S\ref{application} we apply our results to the Minimum Mass
Solar Nebula (MMSN) by modeling thermal 
timescales based on realistic disk properties, and find that the VSI
can operate effectively in the outer parts of the MMSN. We summarize in
\S\ref{summary} with a discussion of caveats and possible directions
for future work.

>>>>

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disks such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, vertical shear is weak in thin disks, 
  so the VSI can be easily stabilized by buoyancy forces. Consequently,  
  the VSI requires a rather short thermal or cooling timescale in
  sub-adabatically (stably) stratified disks.     
  A quantitative examination of this thermodynamic 
  requirement is needed to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of stably stratified accretion disks in which the disk  
  temperature $T$ is relaxed to its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response of the disk by
  varying $\beta$. We find that 
  increasing $\beta$ even slightly away from    
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  is the radial temperature gradient, and 
  $\gamma$ is the adiabatic index. 
  We apply our results  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. In the inner parts of the MMSN, e.g. at $\lesssim 5$AU, 
  we find the VSI would only operate on extremely short radial lengthscales, $\ll 
  O(10^{-2}H)$, where $H$ is the disk scale height. On longer radial 
  lengthscales of $O(10^{-2}H)$ to $O(10^{-1}H)$, we find the MMSN is
  unstable to the VSI from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.  
  % ,and the dominant VSI mode in the MMSN is the `fundamental
  % corrugation mode' which causes the disk column to oscillate
  % vertically.
  We thus suggest the VSI is at least dynamically important in the 
  outer parts of protoplanetary disks at tens of AU. 
\end{abstract}

\section{Introduction}\label{intro}
Understanding the mechanisms of angular momentum transport is the
central problem in the theory of accretion disks 
\citep{lyndenbell74}. The   
magneto-rotational instability in the context of differentially
rotating disks \citep[MRI,][]{balbus91} is now accepted as a robust
pathway to magneto-hydrodynamic (MHD) turbulence, which transports
angular momentum outwards and thus enables mass accretion.

@@

The VSI can be regarded as the disk analog of the
Goldreich-Schubert-Fricke instability 
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
A height-dependence in $\Omega$, or vertical shear, can render
inertial waves unstable. \cite{barker15} gives a simple 
interpretation of the VSI: in an axisymmetric, inviscid, neutrally
stratified disk, the exchange of fluid parcels vertically, while
conserving their specific angular momentum, can result in a new
configuration of lower energy if the distribution of specific angular
momentum varies with $z$. The VSI is characterized by perturbations
with large vertical length-scales  relative to radial length-scales.

Early analysis of the VSI in radially and vertically local 
models of accretion disks \citep{urpin98,urpin03} showed that the VSI
requires a short thermal timescale so that perturbations evolve 
isothermally, unless the disk is neutrally stratified in the vertical
direction. This is because
vertical motions associated with the VSI 
are stabilized by buoyancy forces if the disk is sub-adiabatically
stratified (convectively stable). A sufficiently short thermal
timescale renders buoyancy ineffective, but a quantitative criteria is
desirable. This is the central theme of this work.   
%efficient exchang of heat with surroundings

@@

These studies show that the VSI is generally sensitive to disk
thermodynamics. It is thus important to quantify this dependence, for
example on disk parameters, in order to evaluate the relevance of the
VSI to realistic disks, which one may expect to be stably
stratified and have non-zero thermal timescales.  In order to determine
where in the disk the VSI may operate, and hence lead to hydrodynamic
turbulence, understanding the thermodynamic dependence of the linear VSI is essential. 
% since the transition
% Should a disk
% become turbulent through the VSI, it must first pass through the linear
% instability.

@@

This paper is organized as follows. We first highlight in
\S\ref{vsi_require} the need of a very short thermal timescale for the
VSI to operate in stably stratified disks. There we also preview our
main result --- a quantitative thermodynamic criteria for the VSI ---
and a first application to protoplanetary disks. The remaining sections
are devoted to deriving these results from linear theory.

In \S\ref{setup} we list the governing equations and global
equilibrium disk models under   
consideration. We formulate the linear stability problem in 
\S\ref{linear} and simplify it by making the radially local
approximation. We study the linear problem using two approaches. In
\S\ref{analytical} we present an analytical discussion of vertically
isothermal disks. Our treatment is grossly simplified owing to a
number of approximations made, but closely parallels previous work and 
allows us to highlight the essential effects of vertical shear, and to
derive a characteristic upper limit to the thermal timescale for the
VSI. In \S\ref{numerical} we numerically
solve the linear problem in the radially local approximation, but
relax all other approximations made in our analytical approach.    
We find basic agreement with our analysis, including
the aforementioned characteristic thermal timescale.

In
\S\ref{application} we apply our results to the Minimum Mass
Solar Nebula (MMSN) by modeling thermal 
timescales based on realistic disk properties, and find that the VSI
can operate effectively in the outer parts of the MMSN. We summarize in
\S\ref{summary} with a discussion of caveats and possible directions
for future work.

\section{Thermodynamic requirement for the vertical shear instability}\label{vsi_require}    
In this paper we will find that in a stably stratified disk, the VSI
requires a \emph{very} short thermal timescale $t_c$, much less than the
disk dynamical timescale,
\begin{align}
  t _c \ll \Omega^{-1},  
\end{align}
consistent with numerical experiments \citep{nelson13}. In subsequent
sections, we quantify this requirement more 
precisely. For this preliminary discussion we shall 
consider a vertically isothermal disk with aspect-ratio $\epsilon$,
dimensionless radial temperature gradient $q$ and adiabatic index
$\gamma$.

In a thin disk, the specific angular momentum varies with height as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega^2 }{\p z} \simeq \epsilon q \hat{z} \Omega^2,
\end{align}
where $\hat{z}$ is a normalized vertical co-ordinate. As noted in
\S\ref{intro} this vertical variation is a source of free energy and
thus vertical shear ($\p_z\Omega \neq0$) is destabilizing.

If $\gamma>1$ then the disk is stably stratified with a real vertical
buoyancy frequency $N_z$.  For a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \hat{z}^2
  \Omega^2.  
\end{align}
This provides a stabilizing effect as buoyancy opposes vertical
motion, which is required for the VSI. In fact, if we consider
axisymmetric adiabatic disturbances, then because the 
Solberg-Hoiland criteria \citep{tassoul78} are usually satisified in a
thin disk, the disk would be stable.

However, the stabilizing effect of buoyancy can be mitigated for
non-adiabatic disturbances. Consider a fluid element displaced upwards, away from
the disk midplane. In a stably stratified adiabatic disk, it would
sink because of buoyancy. However, if thermal timescales are short, in
that temperature flucations decay rapidly (i.e. non-adiabatic), then
the fluid element would attain its surrounding density and 
%(assuming an ideal gas and it reaches pressure 
%equilibrium with its surroundings). 
thus buoyancy is rendered ineffective. Qualitatively, a short thermal
timescale $t_c$ reduces the effective (stabilizing) buoyancy force.

How short must $t_c$ be for vertical shear to prevail? We see
from above that the destabilizing effect of vertical shear is rather
weak because $|\p_z\Omega|\propto \epsilon \ll 1$. 
Then even a small amount of buoyancy can significantly stabilize
the VSI. It follows that $t_c$ must be \emph{extremely} short. The
central result of this work is the requirement 
\begin{align}\label{prelim_bcrit}
  (\gamma - 1)t_c < \epsilon |q| \Omega^{-1}
\end{align}   
for the VSI in a vertically isothermal, thin disk. Physically, the  
left-hand side represents an effective buoyancy, while the right-hand
side represents the (weakly)  destabilizing effect of vertical
shear. \emph{In a thin disk the thermal timescale must be shorter than
  the dynamical timescale by $O(\epsilon)\ll 1$ for the VSI to operate.}

% While assumptions and simplifications will be made 
% to derive this,

The simplicity of Eq. \ref{prelim_bcrit} means it can be 
readily applied to evaluate where in a disk the VSI is likely to
operate. For example, we show in Fig. \ref{bcrit_mmsn2d} an estimate
of the thermal timescales in a typical protoplanetary disk model (see 
below). Here, $H$ is the disk scale height and $k_x$ is a radial
wavenumber of the disturbance being considered, thus larger $k_x$
correspond to smaller perturbations. The region outside the red line
satisfies Eq. \ref{prelim_bcrit}.

The derivation of Eq. \ref{prelim_bcrit} in fact
assumes $t_c$ is vertically uniform. While this is not generally true
in the inner disk, we find Eq. \ref{prelim_bcrit} still provides a
useful guide. We see that the VSI is likely to  operate in the outer 
disk at tens of AU, but that smaller scale disturbances may operate at
smaller radii.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.7cm 0cm
  0.73cm]{figures/bcrit_mmsn2d_kx10}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.46cm 0cm
  0.73cm]{figures/bcrit_mmsn2d_kx50}
  \caption{Estimate of the thermal timescale  in a fiducial 
    protoplanetary disk model, the Minimum Mass Solar Nebula. The red
    line corresponds to a characteristic thermal timescale, derived in
    this work, below which the VSI is expected to operate effectively
    (the region outside the line).    
    \label{bcrit_mmsn2d} 
  }
\end{figure}

We now proceed through the formal linear stability problem to arrive
at Eq. \ref{prelim_bcrit} and verify it numerically. The interested 
reader may skip forward \S\ref{application} for the application of our
results to protoplanetary disks.
